### 2020-09-17 text/article_vedrovice.Rmd
percolation analysis

author          : "Petr Pajdla"
    # name          : "Petr Pajdla"
    # affiliation   : "Department of Archaeology and Museology,
    #                  Masaryk University, Brno, Czech Republic"
    # email         : "petr.pajdla@gmail.com"
date            : "`r format(Sys.time(), '%d %B, %Y')`"
abstract        :
  "Spatial patterns in cemeteries and grave goods associations are often explored using various informal methods.
  We adopt computationaly intensive methods utilizing resampling approaches both to spatial analysis and
  distributions of objects from the Early Neolithic cemetery in Vedrovice.

  As a result of the investigation we conclude that...  "

@@

Here we explore the relations between burials and bodies in terms of biological
sex, age, origin, exceptionality of grave goods and location within the
cemetery.
We focus predominantly on investigating non-random patterns in artefact
co-occurences and spatial arragement of the cemetery in relation to the
proposed *levels* of neolithic identity, i.e. sex, age and localness.
Our question is thus how are grave good assemblages related to aforementioned
categories and whether there is a consistent relationship between
these categories forming the *identity*, grave goods and spatial arrangement
of the cemetery.

@@

<!-- site -->
The material analyzed here comes from Vedrovice – Široká u lesa cemetery.

<!-- sample (burials, artefacts) -->
Originally, the burial ground consisted of at least 108 burials.
Some of the burials were disturbed by later settlement and burial pits in the
area of the cemetery, amateur excavations during the 19th century or later
agricultural activities [@podborsky2002].
These burials are excluded from most of the analysis due to missing and/or
incomplete data and only a sample of 78 well-preserved burials is analyzed here.
Most of the data was obtained from the original publication [@podborsky2002],
anthropological determination of sex and age were corrected based on
more recent analysis by @dockalova2008. Carbon, nitrogen and strontium isotope
ratios were collected in @whittle2013 and @richards2008.
Further details on use and function of the artefacts were gathered from
@masclanslatorre2020.
Information on localness of different bodies was adapted from @hrncir2020.

Each burial was originally characterized by 55 variables but only subset of
35 variables was selected for further analysis.
Some of  the original variables are omitted due to their rare occurrence,
for instance spondylus arm ring is present in a single burial only,
thus being useless in establishing relationships between pairs of burials.
Some other variables are removed because they are highly interdependent.

@@

Quantities of various *grave goods* are simplified into a binary matrix indicating
presence or absence of a given trait. 
Only grave goods occuring in more than 5% of the graves are used in the analysis.

@@

To determine non-randomness of various artefact combinations we conduct a
permutation test on the presence/absence data matrix.
The method based on generalized Monte Carlo tests was first
introduced to archaeology by @manly1996 and adapted by @sosna2008.
Our approach is based on both of these works.
Original (observed) co-occurence matrix is permutated $n$ times with constant
row and column totals, i.e. both the count of artifacts in individual graves
and sum of the given type of artifacts across the burials are fixed.

To examine which associations of variables, i.e. artefacts, are infrequent or even not present
and which tend to appear together more often than it would be expected by random chance only
the co-occurences of observed and simulated data
are summarized using two statistics suggested by @manly1996.
The $S$ statistic is a measure of overall deviation between expected and observed
number of artefact occurences while
the $v$ statistic is a measure of difference between co-occurences of
individual artefact with other artifacts
[@manly1996, 475, for details see SI].
To assess relations between groups based on sex, age local origin and exceptionality,
statistics introduced by @sosna2008 are used.

With $n = 999$ permutations of the original data matrix the results obtained
are robust+

## Exceptionality index

Exceptionality index ($EI$) is used in order to induce relative exceptionality
of individual burials based on their attributes.
This approach is independent on a prior selection of attributes of importance
or *prestige*.
It was defined by @demjan2015[367] based on a Root Mean Square (*quadratic mean*) of
PCA factor scores:
$$ EI_{attr_g}=\sqrt{\frac{1}{n}\sum_{i=1}^{n}s_i^2} $$
where $s_i$ are factor scores of individual graves. The mean value is
normalized to range $[0-1]$:

$$ EI_{attr_g}=\frac{EI_{attr_g}}{max~EI_{attr_g}} $$

@@

Another approach considered for establishing neigborhood relations is by
constructing buffer zones around each of the burials.
The question arises how large the buffer zones should be.
To solve this problem we utilise percolation analysis [@maddison2020].
Percolation analysis is a method used to identify clusters of points based
on Euclidean distance.

@maddison2020

The size of the radius is based on the results of nearest neighbors analysis, where
at the given radius there is strong evidence of a clustering process (Fig. \@ref(fig:spat-funs)).
The randomness is again tested using a permutation test similar to the one
described for the neighborhood derived from the graphs.

@@

@@

The quadrat test shows evidence for both random and clustered processes taking place with p values
under the 0.05 alpha value but the $\chi^2$ approximation is inaccurate due to small
expected counts (for the details see supplementary information).
Inconclusive result of the quadrat test is not suprising because as was noted
earlier, we do not expect *first-order* properties to play major roles in
placing of the individual burials.
Results of functions based on nearest neighbors are reported in Fig. \@ref(fig:spat-funs).
*G* function does deviate from the curve expected under independent random process
but does not traverse the boundaries of the envelope.
Thus the cummulative nearest-neighbor distance (*G*) is well approximated by
random process and does not show evidence for spatial clustering nor repelent functions
taking place.
The empty-space function *F* deviates slightly from the area defined by
the envelope at around 6 meters.
This means that the number of nearest neighbours to any empty-space point
at this and larger distances is lower than it would be expected under complete spatial randomness.
On the contrary to only slight deviations in case of *G* and *F*
the *K* and *T* functions show clear deviations from the expected curve and exceed
the boundaries of given envelopes at the distance of around 5 to 6 meters in
case of *K* and slightly more in case of *T* function.
This means that numbers of nearest neighbors at this and larger distances is
significantly higher than we would exect under complete spatial randomness.
The overall point pattern thus shows clear signs of clustering process taking place.

This means that the locations of new burials were not chosen at random
but were affected by the positions of previous burials.
The pattern begins to cluster at the distance of around 5 to 6 meters, meaning
that beginning at this distance, the probability of finding a neighboring burial is
much higher than it would be if the burials were placed randomly.
The community burying their dead at the cemetery of Vedrovice -- Široká u lesa
did so in an organized manner taking into account previous burials at the same place,
meaning their positions were known, probably marked on the surface somehow.
This would imply either non-perishable marks on the surface of the cemetery or their
maintenance from time to time.

Spatial pattern was also explored for differently sexed bodies.

@@

By counting co-occurences of analyzed variables on simulated data matrices and comparing
the distributions of the two studied statistics with the values calculated for
the original dataset we can establish whether the overall pattern of artefacts
distribution in the graves is non-random and further identify artefacts that deviate
from the random pattern.
The value of $S$ statistic is deemed significantly large at the 0.05 alpha level with
$S \approx 14$ and $p \approx xy$ (Fig. \@ref(fig:f-sstat)).
This suggests that there is enough evidence that the overall distribution of
artefacts in the graves is not random.

@@

<!-- ======================================================================= -->
<!-- FIGURES --><!-- FIGURES --><!-- FIGURES --><!-- FIGURES --><!-- FIGURES -->
```{r plan-graphs, fig.cap = "Neighborhood graphs: Delaunay triangulation and Gabriel graph"}
knitr::include_graphics(here("plots/plan_gabriel.pdf"))
```

```{r label-spat-funs}
label_spat_funs <- paste("Functions *G*, *F*, *K* and *T*.",
"*Dashed line* -- theoretical curve under complete spatial randomness, *solid line* -- empirical curve, *gray area* -- envelope.")
```
```{r spat-funs, fig.cap=label_spat_funs}
knitr::include_graphics(here("plots/pointprocess_fun.pdf"))
```

```{r label-nb-sex}
label_nb_sex <- paste("Mean neighbors in a given sex category.",
"*Gray area* -- density estimation for the simulated data, *vertical line* -- observed value.",
"*F* -- female-sexed body, *M* -- male-sexed body, *n. a.* -- non adult, *ind.* -- indetermined sex.")
```
```{r nb-sex, fig.cap=label_nb_sex}
knitr::include_graphics(here("plots/nb_sex.pdf"))
```

```{r f-sstat, fig.cap = "Distribution of simulated and observed S statistic"}
knitr::include_graphics(here("plots/s_statistic.pdf"))
```

```{r f-vstat, fig.cap = "Simulated and observed v statistics for different variables"}
knitr::include_graphics(here("plots", "v_values.pdf"))
```

```{r f-cooc, fig.cap = "Networks of variables frequently present together and exluding each other"}
knitr::include_graphics(here("plots", "cooc_networks.pdf"))
```

<!-- ======================================================================= -->
<!-- TABLES --><!-- TABLES --><!-- TABLES --><!-- TABLES --><!-- TABLES -->
```{r nb-tab-sex}
tab <- read_csv(here("data/temp/nb_sex_p.csv")) %>%
         mutate(relation = str_c(from, to, sep = " - ")) %>%
  select(relation, everything(), -from, -to)

tnames <- c("relation", rep(c("mean", "p"), 3))

knitr::kable(tab, booktabs = TRUE, col.names = tnames, linesep = "", caption = "Results...") %>%
  kableExtra::add_header_above(c(" ", "Buffer" = 2, "Delaunay" = 2, "Gabriel" = 2))
```

>>>>

author          : "Petr Pajdla"
    # name          : "Petr Pajdla"
    # affiliation   : "Department of Archaeology and Museology,
    #                  Masaryk University, Brno, Czech Republic"
    # email         : "petr.pajdla@gmail.com"
date            : "`r format(Sys.time(), '%d %B, %Y')`"
abstract        :
  "Spatial patterns in cemeteries and grave goods associations are often explored using various informal methods.
  We adopt computationaly intensive methods utilizing resampling approaches both to spatial analysis and
  distributions of objects from the Early Neolithic cemetery in Vedrovice.
  As a result of the investigation we conclude that...  "

@@

Here we explore the relations between burials and bodies in terms of biological
sex, age and origin, exceptionality of grave goods and location within the
cemetery.
We focus predominantly on investigating non-random patterns in artefact
co-occurences and spatial arragement of the cemetery in relation to the
proposed *levels* of neolithic identity, i.e. sex, age and localness.
Our question is thus how are grave good assemblages related to aforementioned
categories and whether there is a consistent relationship between
these categories forming the *identity*, grave goods and spatial arrangement
of the cemetery.

@@

<!-- site -->
The material analyzed here comes from Vedrovice – Široká u lesa cemetery.
<!-- sample (burials, artefacts) -->
Originally, the burial ground consisted of at least 108 burials.
Some of the burials were disturbed by later settlement and burial pits in the
area of the cemetery, amateur excavations during the 19th century or later
agricultural activities [@podborsky2002].
These burials are excluded from most of the analysis due to missing and/or
incomplete data and only a sample of 78 well-preserved burials is analyzed here.
Most of the data was obtained from the original publication [@podborsky2002],
anthropological determination of sex and age were corrected based on
more recent analysis by @dockalova2008. Carbon, nitrogen and strontium isotope
ratios were collected in @whittle2013 and @richards2008.
Further details on use and function of the artefacts were gathered from
@masclanslatorre2020.
Information on localness of different bodies was adapted from @hrncir2020.
Local Sr range is defined as two standard deviations from mean of Sr ranges of
juveniles.

Each burial was originally characterized by 55 variables but only subset of
35 variables was selected for further analysis.
Some of  the original variables are omitted due to their rare occurrence,
for instance spondylus arm ring is present in a single burial only,
thus being useless in establishing relationships between pairs of burials.
Only grave goods occuring in more than 5% of the graves are used in the analysis.
<!-- Some other variables are removed because they are highly interdependent.-->

@@

Quantities of various *grave goods* are simplified into a binary matrix indicating
presence or absence of a given trait.

@@

To determine non-randomness of various artefact combinations we conduct a
permutation test on the presence/absence data matrix.
The method based on generalized Monte Carlo tests was first
introduced to archaeology by @manly1996 and adapted by @sosna2008.
Our approach is based on both of these works.
Original co-occurence matrix is permutated *n = 9999* times with constant
row and column totals, i.e. both the count of artifacts in individual graves
and sum of the given type of artifacts across the burials are fixed.

To examine which associations of variables, i.e. artefacts, are infrequent or even not present
and which tend to appear together more often than it would be expected by random chance
the co-occurences of observed and simulated data
are summarized using two statistics suggested by @manly1996.
The *S* statistic is a measure of overall deviation between expected and observed
number of artefact occurences while
the *v* statistic is a measure of difference between co-occurences of
individual artefact with other artifacts
[@manly1996, 475, for details see SI].
To assess relations between groups based on sex, age and origin,
statistics introduced by @sosna2008 are used.

<!-- ## Exceptionality index -->

<!-- Exceptionality index (*EI*) is used in order to induce relative exceptionality -->
<!-- of individual burials based on their attributes. -->
<!-- This approach is independent on a prior selection of attributes of importance -->
<!-- or *prestige*. -->
<!-- It was defined by @demjan2015[367] based on a Root Mean Square (*quadratic mean*) of -->
<!-- PCA factor scores: -->
<!-- $$ EI_{attr_g}=\sqrt{\frac{1}{n}\sum_{i=1}^{n}s_i^2} $$ -->
<!-- where $s_i$ are factor scores of individual graves. The mean value is -->
<!-- normalized to range $[0-1]$: -->

<!-- $$ EI_{attr_g}=\frac{EI_{attr_g}}{max~EI_{attr_g}} $$ -->

@@

<!-- Percolation analysis -->
Another approach considered for establishing neigborhood relations is by
constructing buffer zones around each of the burials.
The question arises how large the buffer zones should be.
To avoid this problem we utilise percolation analysis.
Percolation analysis is a method used to identify clusters of points based
on spatial separation measured by Euclidean distance [@maddison2020, 269].
The introduction of percolation analysis to archaeology is relatively recent,
nevertheless its effectivness and usefulness in studies with varied scales
is already proven [@maddison2020].
Percolation analysis is based on city clustering algorithm [@rozenfeld2008] and
its generalization utilizing Euclidean distance [@rozenfeld2011].
Nodes being in a given distance threshold (radius) are assigned into a cluster.
This process is repeated for a set of thresholds while recording maximum
cluster sizes and other statistics.
This allows to identify distances at which clusters exhibit rapid grow (absorbing
large number of nodes) and distances at which the clusters remain relatively stable.
For more details refer to @maddison2020 and associated R package [@schmidt2020].

<!-- The size of the radius is based on the results of nearest neighbors analysis, where
at the given radius there is strong evidence of a clustering process (Fig. \@ref(fig:spat-funs)).
The randomness is again tested using a permutation test similar to the one
described for the neighborhood derived from the graphs.-->

@@

## Artifact combinations

By counting co-occurences of analyzed variables on simulated data matrices and comparing
the distributions of the two studied statistics with the values calculated for
the original dataset we can establish whether the overall pattern of artefacts
distribution in the graves is non-random and further identify artefacts that deviate
from the random pattern.
The overall deviation from expected co-occurences, i.e. the *S* statistic
[@manly1996, 475] is significantly large at the 0.05 alpha level.
This suggests that there is strong evidence that the overall distribution of
artefacts in different graves is not random (see Fig. \@ref(fig:sstat)).

Examining distributions of individual artefacts it is found that at the 0.05
significance level only adzes, grinding tools, ceramic bowls and spondylus beads
are distributed randomly.
Close to the 0.05 alpha level are distributions of bone tools and ceramic bottles
(Fig. \@ref(fig:vstat)).
Among the grave goods that are found to be randomly distributed in the graves
across the burial ground are ochre pigment, lithics, pebbles, globular pots and
special types of pottery vessel shapes and spondylus pendants and *buckles*.
The relationships between artefact types frequently occuring and missing together
in graves are visualized in Fig. (\@ref(fig:coocnets)) with
clusters highlighted using fast-greedy algorithm [@newman2004] and strength of the
given association is expressed by the thickness of the edge connecting the nodes.
Artifacts most frequently present with other artifacts are with 5 links grinding tools and
ceramic bowls and with 4 links adzes, non-local lithic tools, ochre pigment and
spondylus beads.
The association is strongest between spondylus beads and (a) ceramic bottles and
(b) pigment; and adzes and (a) non-local lithics and (b) bone tools.
Artifacts that are frequently absent in the graves together are with 6 links
local lithic tools, globular pots and ceramic bowls and spondylus buckels with
5 and 4 links respectively.
The strongest antagonist relationships are between globular pots and (a) bowls and
(b) adzes.

@@

The quadrat test shows evidence for both random and clustered processes taking place with p values
under the 0.05 alpha level but the $\chi^2$ approximation is inaccurate due to small
expected counts (for the details see supplementary information).
Inconclusive result of the quadrat test is not suprising because as was noted
earlier, we do not expect *first-order* properties to play major roles in
placing of the individual burials.
Results of functions based on nearest neighbors are reported in Fig. \@ref(fig:spat-funs).
*G* function does deviate from the curve expected under independent random process
but does not traverse the boundaries of the envelope.
Thus the cummulative nearest-neighbor distance (*G*) is well approximated by
random process and does not show evidence for spatial clustering nor repelent functions
taking place.
The empty-space function *F* deviates slightly from the area defined by
the envelope at around 6 meters.
This means that the number of nearest neighbours to any empty-space point
at this and larger distances is lower than it would be expected under complete spatial randomness.
On the contrary to only slight deviations in case of *G* and *F*
the *K* and *T* functions show clear deviations from the expected curve and exceed
the boundaries of given envelopes at the distance of around 5 to 6 meters.
<!-- in case of *K* and slightly more in case of *T* function.-->
This means that numbers of nearest neighbors at this and larger distances is
significantly higher than we would exect under complete spatial randomness.
The overall point pattern thus shows clear signs of clustering process taking place.

This means that the locations of new burials were not chosen at random
but were affected by the positions of previous burials.
The pattern begins to cluster at the distance of around 5 to 6 meters, meaning
that beginning at this distance, the probability of finding a neighboring burial is
much higher than it would be if the burials were placed randomly.
The community burying their dead at the cemetery of Vedrovice -- Široká u lesa
did so in an organized manner taking into account previous burials at the same place,
meaning their positions were known, probably marked on the surface somehow.
This would imply existence of non-perishable marks on the surface of the cemetery and/or their
maintenance from time to time.

Results of percolation analysis reveal more detailed view of the clustering process
taking place.
The curve in Fig. \@ref(fig:perc-distance) reveals that most of the growth into
a single cluster happens at distances between approximately 4 and 6 meters.
The distances where relative stability of clusters emerges are labeled.
The clusters at these distances are shown in Fig. \@ref(fig:perc-plan).

@@

@@

<!-- ======================================================================= -->
<!-- FIGURES --><!-- FIGURES --><!-- FIGURES --><!-- FIGURES --><!-- FIGURES -->

```{r sstat, fig.cap = "Distribution of simulated and observed *S* statistic"}
knitr::include_graphics(here("plots/s_statistic.pdf"))
```

```{r vstat, fig.cap = "Simulated and observed *V* statistics for different variables"}
knitr::include_graphics(here("plots", "v_values.pdf"))
```

```{r coocnets, fig.cap = "Networks of variables frequently present together and exluding each other"}
knitr::include_graphics(here("plots", "cooc_networks.pdf"))
```

<!-- ```{r plan-graphs, fig.cap = "Neighborhood graphs: Delaunay triangulation and Gabriel graph"} -->
<!-- knitr::include_graphics(here("plots/plan_gabriel.pdf")) -->
<!-- ``` -->

<!-- ```{r label-spat-funs} -->
<!-- label_spat_funs <- paste("Functions *G*, *F*, *K* and *T*.", -->
<!-- "*Dashed line* -- theoretical curve under complete spatial randomness, *solid line* -- empirical curve, *gray area* -- envelope.") -->
<!-- ``` -->
<!-- ```{r spat-funs, fig.cap=label_spat_funs} -->
<!-- knitr::include_graphics(here("plots/pointprocess_fun.pdf")) -->
<!-- ``` -->

<!-- ```{r label-nb-sex} -->
<!-- label_nb_sex <- paste("Mean neighbors in a given sex category.", -->
<!-- "*Gray area* -- density estimation for the simulated data, *vertical line* -- observed value.", -->
<!-- "*F* -- female-sexed body, *M* -- male-sexed body, *n. a.* -- non adult, *ind.* -- indetermined sex.") -->
<!-- ``` -->
<!-- ```{r nb-sex, fig.cap=label_nb_sex} -->
<!-- knitr::include_graphics(here("plots/nb_sex.pdf")) -->
<!-- ``` -->

<!-- ======================================================================= -->
<!-- TABLES --><!-- TABLES --><!-- TABLES --><!-- TABLES --><!-- TABLES -->
<!-- ```{r nb-tab-sex} -->
<!-- tab <- read_csv(here("data/temp/nb_sex_p.csv")) %>% -->
<!--          mutate(relation = str_c(from, to, sep = " - ")) %>% -->
<!--   select(relation, everything(), -from, -to) -->

<!-- tnames <- c("relation", rep(c("mean", "p"), 3)) -->

<!-- knitr::kable(tab, booktabs = TRUE, col.names = tnames, linesep = "", caption = "Results...") %>% -->
<!--   kableExtra::add_header_above(c(" ", "Buffer" = 2, "Delaunay" = 2, "Gabriel" = 2)) -->
<!-- ``` -->
